\newpage

\begin{dedicatoria}
    \vspace*{\fill}
    %\noindent
    \hfill
    \begin{minipage}{.6\textwidth}
        I dedicate this dissertation to my grandfather	Adenir (in memoriam), who always encouraged me to study, even
        when he did not understand it.
    \end{minipage}
\end{dedicatoria}
 
\begin{agradecimentos}
	I thank everyone who contributed along this journey, in particular:

    My parents Fabiana and André, who have always been by my side since my
    inception and supported my choices, and my family, who understood the distance from my hometown.

    My professors of the Applied Mathematics course at Fundação Getulio Vargas for technical contributions throughout the course. In special, for my Scientific Initiation professor, Maria Soledad Aronna, who helped me start the scientific journey. 

    My friends and math colleagues for emotional support and discussion, including Allan, Cristiana, Giovani, João, Lucas, and others.

    Leonardo Soares Bastos and Camila Mattos dos Santos for helpful discussions about Respondent-driven sampling.

    FGV's Center for the Development of Mathematics and Science, who believed
    in my work and financially and emotionally assisted me along the course.
    
    National Council for Scientific and Technological Development (CNPq) for
    granting me research funding.

    My advisor Luiz Max Carvalho for agreeing to conduct my research work and
    helping me with all the steps for the work conclusion, including English
    enhancement, reference indications, discussions about results and
    suggestions for improvements. I also thank his quick email responses which
    accelerated the learning process. 

    Finally, I thank scientific researchers and science communicators for enlightening through knowledge in difficult times.
\end{agradecimentos}

\begin{epigrafe}
\vspace*{\fill}

\begin{flushright}
    \hspace{7.5cm}
    \textit{
        ``One of the pervasive risks that we face in the \\ information age is
        that even if the amount of \\ knowledge in the world is increasing,
        the gap \\ between what we know and what we think \\ we know may be
        widening.''} \\
        \textit{Nate Silver}
\end{flushright}
\end{epigrafe}